Recurrence spike spectra

### 2022-06-28 Manuscript/inter_spike_manuscript.tex
Laskar results: no success... :/

%
\smartqed  % flush right qed marks, e.g. at end of proof
%
\usepackage{graphicx}
%
% \usepackage{mathptmx}      % use Times fonts if available on your TeX system
%
% insert here the call for the packages your document requires
%\usepackage{latexsym}
\usepackage{dcolumn}% Align table columns on decimal point
\usepackage{bm}% bold math
\setlength{\marginparwidth}{2cm}
\usepackage{changes}
\usepackage{bbm}
%\usepackage[mathlines]{lineno}% Enable numbering of text and display math
%\linenumbers\relax % Commence numbering lines
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{mathptmx}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{statmath}
\usepackage[numbers]{natbib}
\usepackage{hyperref}
\usepackage{enumitem}
\usepackage{mathtools}
\DeclarePairedDelimiter\ceil{\lceil}{\rceil}
%
% please place your own definitions here and don't use \def but
% \newcommand{}{}
%
% Insert the name of "your journal" with
\journalname{Nonlinear Dynamics}
%
\begin{document}

@@

\subsection{Evolutionary spike spectrogram of CENOGRID isotope time series}

%To demonstrate XXXXXX we applied it to the recently published CENOzoic Global Reference
%benthic foraminifer carbon and oxygen Isotope Dataset (CENOGRID) \cite{Westerhold2020}. 
%The temperature-dependent fractionation of carbon and oxygen isotopes in benthic foraminifera
%is an important means to reconstruct past global temperatures and environmental conditions.
%Moreover, the Cenozoic is interesting, because it provides an analogue of future greenhouse 
%climate and how and which regime shifts in large-scale atmospheric and ocean circulation 
%can be expected in the future warming climate. The dataset consists of a detrended $\delta^{18}$O and a detrended $\delta^{13}$C isotope record with a total length
%of $N=13,421$ samples and a sampling period of $\Delta t= 5,000 \text{yrs}$ (Figure \ref{fig_time_series_cenogrid} in Appendix \ref{sec_appendix_cenogrid}). Here we made predictions 
%on the $\delta^{13}$C isotope record.
%We again utilize the Cenozoic Global Reference benthic foraminifer carbon and oxygen Isotope Dataset (CENOGRID) \cite{Westerhold2020}. The time series 
%have a total length of $N=13,421$ samples and a sampling period of 
%$\Delta t= 5,000 \text{yrs}$. The records and their corresponding RPs are shown in 
%Figure~\ref{fig_tauRR_cenogrid_RPs} together with remarkable climate transitions in the Cenozoic (detrended time series in Fig.~\ref{fig_time_series_cenogrid} in 
%Appendix~\ref{sec_appendix_cenogrid}). Among other things, the large-scale patterns in the RPs shown were used by \textcite{Westerhold2020} to classify major climate 
%states: Hothouse, Warmhouse, Coldhouse and Icehouse. 
%\begin{figure}[h!]
% \includegraphics[width=\textwidth]{./figures/chapter06_tauRR/fig_tauRR_cenogrid_RPs}
% \caption{\textbf{A} $\delta^{13}$C and \textbf{B} $\delta^{18}$O isotope records and their corresponding RPs (\textbf{C+D}), obtained from the non-embedded time 
% series under a fixed recurrence threshold corresponding to 8\% global RR. These plots were used by \textcite{Westerhold2020} to classify the major climate states 
% Hothouse, Warmhouse, Cold- and Icehouse.
%\label{fig_tauRR_cenogrid_RPs}

@@

A novel type of powerspectrum, the \textit{inter spike spectrum}, has been proposed. The method decomposes any arbitrary signal into basis functions which consist of (lagged) Dirac 
combs (DC) of different inter-spike period. The loading for each period is obtained by a regularized regression, which promotes sparsity in its solution (we chose LASSO). Since there are 
$M = \sum_{i=1}^{N}i$ basis functions for a signal of length $N$ the regression can get computationally intensive for $N>1,000$. When plotting the computed loadings as a function 
of the frequency (or period) the inter spike spectrum is obtained. This novel powerspectrum is superior to an ordinary FFT-based powerspectrum, when the signal has a spike-train-like 
appearance. The invention of the proposed method has been motivated by the idea of transforming $\tau$-recurrence rate signals ($\tau$-RR's) into their frequency domain. 
This general idea \cite{Zbilut2008} allows for a frequency analysis of high dimensional systems, because the RP is a representation of the system's state space trajectory.   
The $\tau$-RR of a recurrence plot (RP) usually has a spiky shape, especially for map-like data, and the inter spike spectrum can reliably reveal the system's dominant frequencies, 
which is not possible when Fourier transforming the $\tau$-RR or the underlying signal itself. Since the position of the peaks 
in the $\tau$-RR are not sensitive to noise, the corresponding inter spike spectrum also yields robust results in the presence of noise. We have successfully used the idea 
of transforming the $\tau$-RR for the detection of bifurcations in the Logistic map. By constructing appropriate surrogates of the inter spike spectra, and thus a null model, 
the number of significant peaks in the inter spike spectrum correlated well with the positive Lyapunov exponent. This measure was also able to resolve period-doubling bifurcations.

We could think of a broad range of applications of the proposed idea. The inter spike spectrum itself can serve as a valuable tool for the analysis of any sort of 
spike-train-like data. On the other hand, the inter spike spectrum of the $\tau$-RR of a signal can serve as a generalized, nonlinear frequency analysis tool for complex systems. 
When there is only a subset of state variables available, the state space has to be reconstructed as a pre-processing step. Recent findings \cite{Kraemer2021} show that 
this reconstruction process can be reliably automated and applied to multivariate data as well. This would allow for a ``running window'' approach, in order to detect transitions. 
Due to the mentioned computational constraints of our proposed method, a window size $w\leq 1,000$ would possibly suffice for most data, especially when it is map-like, i.e., not 
highly sampled.

\begin{acknowledgements}
This work has been financially supported by the German Research Foundation (DFG projects MA4759/8 and MA4759/9). All computations have been carried out in the Julia language 
and made use of the packages \textit{DynamicalSystems.jl} \cite{datseris2018} and \textit{DifferentialEquations.jl} \cite{rackauckas2017}.
\end{acknowledgements}

@@

>>>>

%
\smartqed  % flush right qed marks, e.g. at end of proof
%
\usepackage{graphicx}
%
% \usepackage{mathptmx}      % use Times fonts if available on your TeX system
%
% insert here the call for the packages your document requires
%\usepackage{latexsym}
\usepackage{dcolumn}% Align table columns on decimal point
\usepackage{bm}% bold math
\setlength{\marginparwidth}{2cm}
\usepackage{changes}
\usepackage{bbm}
%\usepackage[mathlines]{lineno}% Enable numbering of text and display math
%\linenumbers\relax % Commence numbering lines
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{mathptmx}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{statmath}
\usepackage[numbers]{natbib}
\usepackage{hyperref}
\usepackage{enumitem}
\usepackage{mathtools}
\DeclarePairedDelimiter\ceil{\lceil}{\rceil}
%
% please place your own definitions here and don't use \def but
\newcommand{\mytilde}{\raise.17ex\hbox{$\scriptstyle\mathtt{\sim}$}}
%
% Insert the name of "your journal" with
\journalname{Nonlinear Dynamics}
%
\begin{document}

@@

\subsection{Evolutionary spike spectrogram of Earth's orbit time series}

To demonstrate the usefulness of the proposed inter spike spectrum, we apply it to the freely available eccentricity time series of 
\citet{Laskar2011}. \todo[inline]{blabla what is this used for usually? etc.} We use a time series which covers the past \mytilde $67$ Mio. years, which a total length
of $N=13,421$ samples and a sampling period of $\Delta t= 5,000$.

@@

A novel type of powerspectrum, the \textit{inter spike spectrum}, has been proposed. The method decomposes any arbitrary signal into basis functions which consist of (lagged) Dirac 
combs (DC) of different inter-spike period. The loading for each period is obtained by a regularized regression, which promotes sparsity in its solution (we chose LASSO or a sequentially 
thresholded least squares regression STLS). Since there are 
$M = \sum_{i=1}^{\ceil*{N/2}+1}i$ basis functions for a signal of length $N$ the regression can get computationally intensive for $N>1,000$. When plotting the computed loadings as a function 
of the period (or frequency) the inter spike spectrum is obtained. This novel powerspectrum is superior to an ordinary FFT-based powerspectrum, when the signal has a spike-train-like 
appearance. Due to the sparse regression underlying the method, there is no unique inverse of the transformation and the regularization parameter plays a crucial role and determines the 
appearance of the obtained inter spike spectrum. Moreover, similar to the Nyquist frequency barrier in the Fourier Transform which sets a lower bound for the corresponding wave period, here 
the maximum considered inter spike period is bounded by $T_{\text{is}}^{\text{max}} = \ceil*{N/2}+1$.

The invention of the proposed method has been motivated by the idea of transforming $\tau$-recurrence rate signals ($\tau$-RR's) into their frequency domain. 
This general idea \cite{Zbilut2008} allows for a frequency analysis of high dimensional systems, because the RP is a representation of the system's state space trajectory.   
The $\tau$-RR of a recurrence plot (RP) usually has a spiky shape, especially for map-like data, and the inter spike spectrum can reliably reveal the system's dominant frequencies, 
which is not possible when Fourier transforming the $\tau$-RR or the underlying signal itself. Since the position of the peaks 
in the $\tau$-RR are not sensitive to noise, the corresponding inter spike spectrum also yields robust results in the presence of noise. We have successfully used the idea 
of transforming the $\tau$-RR for the detection of bifurcations in the Logistic map. By constructing appropriate surrogates of the inter spike spectra, and thus a null model, 
the number of significant peaks in the inter spike spectrum correlated well with the positive Lyapunov exponent. This measure was also able to resolve period-doubling bifurcations. 
\todo[inline]{Here a quick summary of the other applications.}

We could think of a broad range of applications of the proposed idea. The inter spike spectrum itself can serve as a valuable tool for the analysis of any sort of 
spike-train-like data. On the other hand, the inter spike spectrum of the $\tau$-RR of a signal can serve as a generalized, nonlinear frequency analysis tool for complex systems. 
When there is only a subset of state variables available, the state space has to be reconstructed as a pre-processing step. Recent findings \cite{Kraemer2021,Kraemer2022} show that 
this reconstruction process can be reliably automated and applied to multivariate data as well. This would allow for a ``running window'' approach, in order to detect transitions. 
Due to the mentioned computational constraints of our proposed method, a window size $w\leq 1,000$ would possibly suffice for most data, especially when it is map-like, i.e., not 
highly sampled.

\begin{acknowledgements}
This work has been financially supported by the German Research Foundation (DFG projects MA4759/8 and MA4759/9). All computations have been carried out in MATLAB\textsuperscript{\textregistered} 
and the Julia language and made use of the packages \textit{DynamicalSystems.jl} \cite{datseris2018} and \textit{DifferentialEquations.jl} \cite{rackauckas2017}.
\end{acknowledgements}

@@

\section*{Code availability}

The study that we present here is available as a fully reproducible code base
**Repository will be published and cited here, when accepted** and the method
will be available in the Julia language **Package will be published and cited here, 
when accepted** and as a MATLAB\textsuperscript{\textregistered} toolbox 
**Toolbox will be published and cited here, when accepted**.
